\subsubsection{Marketplace Structure}
The platform is a consumer-to-consumer (C2C) online marketplace. It facilitates the sale of secondhand goods, primarily apparel, alongside home goods and electronics. The business model is commission-based; the firm collects a fee for each transaction. For sales of \$15 or more, the commission is 20% of the final price. For sales below \$15, the fee is a flat \$2.95. This structure aligns platform revenue with the gross merchandise value (GMV) transacted between users.

\subsubsection{Social Commerce Mechanics}
The platform integrates social networking features into the e-commerce experience. User actions such as following, liking, and sharing listings are core to the product. A seller's listing visibility is influenced by their level of engagement within the community. This design establishes a system where seller effort, in the form of social participation, is a direct input for organic discovery. Features such as themed, real-time ``Posh Parties'' and live-selling ``Posh Shows'' further embed social interaction into the transaction process.

\subsubsection{Post-Acquisition Strategic Realignment}
In January 2023, Naver Corporation acquired the firm, signaling a strategic realignment. This shift occurred as the firm consolidated its operations to North America, having ceased activities in its UK, Australian, and Indian markets. The new strategy focused on developing monetization channels beyond the commission-based model, which accounted for approximately 96% of total revenue in 2021.

This led to the introduction of ``Promoted Closet,'' a platform-native advertising tool for sellers in the United States, creating a dual revenue model. The service is an automated system where sellers set a weekly budget to have their listings promoted in high-visibility placements, such as search results and brand pages. The platform's algorithm surfaces these listings when a sale is deemed probable. It operates on a cost-per-click (CPC) basis, with a variable CPC and weekly billing. A seller's budget size directly impacts listing visibility; the platform provides a recommended budget based on closet size and advises sellers to increase it when listing more items. Budget changes take effect at the start of the next weekly cycle. The tool is available to all sellers without minimum requirements. A 28-day attribution window is used; a sale of any item from a seller's closet, including those made via offers, is attributed to the campaign if the buyer clicked any of their promoted listings. The seller dashboard provides metrics such as Return on Ad Spend (ROAS) and ranks ``Top Promoted Listings.'' To encourage adoption, the platform offers free trials that automatically convert to paid weekly subscriptions. This development signals a fundamental change in the marketplace's economic design, introducing a mechanism for sellers to convert capital, in addition to effort, into visibility.
